Supplementary Materials

### 2022-12-24 manuscript/supplementary.tex
paper updates

\begin{align*}
  n_T &= \frac{n(n-1)}{2} \\
  n_S &= {c_1 \choose 2} + {c_2 \choose 2} \\
  n_D &= n_T - n_S
\end{align*}
%
which would be stored in row $\mathbf{Y}_l = [n_S, n_D]$.

@@

\section{}

>>>>

\begin{align*}
  n_T &= \frac{n(n-1)}{2} \\
  n_S &= {c_1 \choose 2} + {c_2 \choose 2} \\
  n_D &= n_T - n_S
\end{align*}
%
which would be stored in row $\mathbf{Y}_l = [n_S, n_D]$. Note that the
per-site $\mathbf{Y}$ handles non-polymorphic sites and missing data.
Non-polymorphic sites have $n_S = {n \choose 2}$ and $n_D = 0$, and missing
data has $n_S = n_D = 0$.

@@

\section*{Likelihood}

Our model is essentially a generalized nonlinear model with a binomial link
function. This is the form used by \textcite{Elyashiv2016-vt} and
\textcite{Murphy2022-sj}. These models fit the observed number of pairwise
nucleotide differences in genomic windows to the expected pairwise diversity
under some evolutionary model. We only consider BGS models here, so the mean
function for position $x$ is the product of the proportion by which BGS reduces
diversity at $x$, $B(x)$, and genome-wide neutral diversity $\pi_0$,

\begin{align}
  \pi(x) = B(x, \Phi) \pi_0
\end{align}

$\Phi$ is the set of BGS parameters (i.e. the DFE for each feature type and
mutation rate), and $\pi_0$ is determined by the drift-effective population
size, set by only reproductive and reproductive processes. Then the ob

fit a nonlinear theoretic equation to
observed count data summarized at some pre-specified scale.

putatively neutral site, the observed number of pairwise differences at site
$i$ is modeled as,

\begin{align}
  n_{D,i} \sim \text{Binom}(\pi_i, n_{T,i})
\end{align}

use observed count data summarized at
some pre-specified scale

\begin{align}
  \log\mathcal{L}(\theta) = \sum_{v \in \mathcal{V}} \sum_{i \ne j \in \mathcal{S}} \log(P(O_{i,j}(v) | \theta))
\end{align}

where $\mathcal{V}$ is the set of putatively neutral sites, $\mathcal{S}$ is
the set of samples, and $\theta$ are the BGS parameters. The indicator variable
$O_{i,j}(v)$ is 1 if samples $i$ and $j$ are different at site $v$, and zero
otherwise. Thus as they specify in the paper,

\begin{align}
  P(O_{i,j}(v) | \theta) = 
    \begin{cases}
      \pi(v | \theta), & O_{i,j}(v) = 1 \\
      1-\pi(v | \theta), & O_{i,j}(v) = 0 \\
    \end{cases}.
\end{align}

The total size of the set of samples $\mathcal{S}$ is $n_\mathcal{S} =
|\mathcal{S}|$. Assuming all sites are biallelic, we can simplify the inner
summation by counting the number of possible same and different pairwise
combinations. If a site $v$'s vector of allele counts is $[c_1, c_2]$, the
total number of pairwise combinations with the same alleles is

\begin{align}
  n_s(v) = {c_1 \choose 2} + {c_2 \choose 2}
\end{align}

and the number of different pairwise combinations is

\begin{align}
  n_d(v) = n_T - n_s(v)
\end{align}

where $n_T = n_\mathcal{S} (n_\mathcal{S} - 1) / 2$  is the total number of
pairwise combinations across the sample set $\mathcal{S}$. Note that these
site-specific counts allow us to use allelic counts directly, and can vary
across sites. Our log-likelihood is then,

\begin{align}
  \ell(\theta) &= \sum_{v \in \mathcal{V}} \left[\log(\pi(v | \theta)) n_\text{D}(v) + \log(1-\pi(v | \theta)) n_\text{S}(v)\right].
\end{align}

In practice, we calculate these values across bins. For each bin, we treat
$\pi(v | \theta)$ as fixed, assuming that at this scale, the variation in
expected diversity across sites is minimal. For a particular chromosome, we
have two classes of sites: those included in the diversity calculation and
those ignored. The former sites are all putatively neutral and have reliably
called genotypes, and the other sites are possibly non-neutral or do have
reliably called genotypes. The total log-likelihood is the sum of bin likelihoods,
$\ell(b)$

\begin{align}
   \ell(\theta) =  \sum_b \ell(b | \theta)
\end{align}

The likelihood within a bin is then,

\begin{align}
  \ell(b | \theta)  &= \log(\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{V}_b} n_D(v) + \log(1-\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{V}_b} n_S(v)  \\
                               &= \log(\bar{\pi}(b | \theta)) Y_D(b) + \log(1-\bar{\pi}(b | \theta)) Y_S(b)
\end{align}
%
where the two sum terms as $Y_D(b)$ and $Y_S(b)$ are data reductions at the bin
level.

If the data are such that only polymorphic sites are considered, we can adapt
this by partitioning the set $\mathcal{V}$ of neutral sites into polymorphic
($\mathcal{P}$) and fixed sites ($\mathcal{F}$), i.e. $\mathcal{V} =
\mathcal{P} \cup \mathcal{F}$ and $\mathcal{P} \cap \mathcal{F} = \varnothing$.
For all $v \in \mathcal{F}$, $n_d(v) = 0$ and $n_s(v) = n_T$.

Then,

\begin{align}
  \ell(\theta) &= \sum_{v \in \mathcal{V}} \left[\log(\pi(v | \theta)) n_\text{D}(v) + \log(1-\pi(v | \theta)) n_\text{S}(v)\right] \\
                  &= \sum_{v \in \mathcal{P}} \left[\log(\pi(v | \theta)) n_\text{D}(v) + \log(1-\pi(v | \theta)) n_\text{S}(v)\right] + \sum_{v \in \mathcal{F}} \log(1-\pi(v | \theta)) n_T(v)  \\
\end{align}

\begin{align}
  \ell(b | \theta)  &= \log(\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{P}_b} n_\text{D}(v) + \log(1-\bar{\pi}(b | \theta)) \left(\sum_{v \in \mathcal{P}_b} n_\text{S}(v) +  \sum_{v \in \mathcal{F}_b} n_T(v)  \right).
\end{align}

Note that if we assume that the total number of combinations at each fixed site
is constant, e.g. $n_T = n_T(v)$ for all $v$, then we can use $\sum_v n_T(v) =
n_T |\mathcal{F}_b|$.

\
\section*{The }

The core parts of our likelihood are,

\begin{align}
  \ell(\theta) &=  \sum_b [\log(\bar{\pi}(b | \theta)) Y_D(b) + \log(1-\bar{\pi}(b | \theta)) Y_S(b)]
\end{align}

where,

\begin{align}
  \bar{\pi}(b |\theta) = \pi_0(b) \bar{B}(b | \theta).
\end{align}

Here, $\bar{B}(b | \theta)$ is the predicted reduction in diversity due to BGS
in window $b$, given background selection parameters $\theta$. In practice,
this is site-specific. We can write the reduction at any neutral site $v$ in
the genome as the product of $B$s across all segments,

\begin{align}
  B(v | \theta) = \exp\left(- \sum_g \int f(\mu(\mathcal{A}(g)), s, S_g) w(s|\mathcal{A}(g)) ds \right)
\end{align}

where $S_g$ is exogenous genomic data about the segment, $S_g = \{L_g, r_g,
\rho(|v-p_g|)\}$, where $L_g$ is the segment's length, $r_g$ is the
recombination rate per basepair in the segment, and $\rho(|v-p_g|)$ is the
recombination distance between the focal site $v$ and the segment position
$p_g$ (we approximate, and use the nearest end position to the neutral site).
Additionally, $w(s|\mathcal{A}(g))$ is the distribution of selection
coefficients for segment $g$, if segment $g$ is a member of annotation class
$\mathcal{A}(g)$.

We can think about the DFE as the conditional distribution of a particular
selection coefficient given a mutation occurs, for a particular annotation
class. The BGS function $f(\cdot)$ only depends on $\mu$ through the
introduction of deleterious alleles with selection coefficient $s$ at rate
$\omega(s|\mathcal{A}(g)) = \mu(\mathcal{A}(g)) w(s | \mathcal{A}(g))$. Thus,
we can write,

\begin{align}
  B(v | \theta) = \exp\left(- \sum_g \int f(\omega(s|\mathcal{A}(g)), s, S_g) ds \right)
\end{align}

which we can discretize as,

\begin{align}
  B(v | \theta) = \exp\left(- \sum_g \sum_s f(\omega(s|\mathcal{A}(g)), s, S_g) \right).
\end{align}

Next, note that there are a finite number of annotation classes, $\mathcal{A}
\to \{a_1, a_2, \ldots, a_k\}$, so we can further partition this as

\begin{align}
  B(v | \theta) = \exp\left(- \sum_{\{g \;:\; \mathcal{A}(g) = a_1\}} \sum_s f(\omega(s| a_1), s, S_g) + \sum_{\{g \;:\; \mathcal{A}(g) = a_2\}} \sum_s f(\omega(s| a_2), s, S_g) + \ldots  \right)
\end{align}

Let us define the $d_\omega \times d_s \times d_g$ multidimensional array
$\mathbf{F}$, and the $d_g \times d_a$ feature classification matrix
$\mathbf{A}$.

\section{Calculation of B Maps}

Each B map is determined by the annotation track of putatively conserved
segments, which includes (1) the segment positions and lengths, (2) the type of
feature of each segment (e.g.  exons, UTRs, phastcons conserved elements,
etc.), as well as (3) the recombination map, and (3) the deleterious selection
coefficient $s(i)$ and mutation rate at which these deleterious enter the
population for feature type $i$.

The B maps are computed at fixed, evenly spaced sites.

\section{Ratchet Rate Prediction}

\begin{align}
  R = \int 
\end{align}
